Vagueness, Noise, and Signaling

### 2015-01-26 paper/version_02/paper.tex
Renamed section 5.2.
paper revision 2 some (almost) final edits

\begin{abstract}
  Signaling games are popular models for studying the evolution of
  meaning, but typical approaches do not incorporate vagueness as a
  feature of successful signaling.  Complementing recent like-minded
  models, we introduce the replicator diffusion dynamic as a special
  case of the replicator mutator dynamic that combines evolutionary
  pressure on optimal signal use with stochastic noise in the
  processing of similar stimuli. Applying this new dynamic to a
  generalization of Lewis' signaling games, we show that stochastic
  imprecision leads to vague, yet by-and-large communicative efficient
  signal use, and, moreover, that it unifies evolutionary outcomes and helps
  avoiding suboptimal categorization.
\end{abstract}

@@

An alternative approach is taken by
\citet{FrankeJager2010:Vagueness-Signa} and
\citet{OConnor2013:The-Evolution-o} who show how the above desiderata
can be met by evolving strategies in sim-max games. The approach we
take here is similar in spirit, but different in relevant detail. We
argue here that \citeauthor{FrankeJager2010:Vagueness-Signa}'s
approach gives rise to problematic predictions under certain conditions
which are avoided in our approach. However, the two are not necessarily
competing explanations since they are motivated differently and it is
certainly possible that both the mechanisms proposed by them, as the
ones proposed here play a role in the whole phenomenon.
Regarding the work by \citeauthor{OConnor2013:The-Evolution-o}, while she
gives an agent-level learning dynamic under which vague signaling arises,
we complement the picture by giving a population-level dynamic that
abstracts from the concrete behavior of signalers. A more in-depth
comparison is deferred until Section~\ref{sec:discussion}. Let us
first look into our own proposal in more detail.

@@

The most natural interpretation of this sequential definition is that
the \rdd captures the most likely path of development of aggregate
behavior in a society, in which behavioral
changes occur independently at each choice point, as described by the
\rd for behavioral strategies. Whatever strategy an agent does
actually hold, its realization is perturbed by imperfect
discriminability of states in such a way that the expected overall
aggregate behavior is the diffusion of the strategies that the agents
would play in a noise-free world. Any agent-level interpretation of
the \rd in behavioral strategies seems compatible with this
interpretation.

The sequential definition highlights that other (discrete-time)
evolutionary dynamics can be subjected to state-confusability in an
analogous fashion. It would be entirely straightforward, for instance,
to perturb a best-response dynamic in case we wanted to model
cognizing, rational agents with a noisy perceptual system. Our choice
of the \rd as a carrier dynamic is motivated by its generality and
austerity, being compatible with a variety of implicit agent-level
behavior, without presupposing extravagant higher cognitive abilities
\citep[c.f.][]{Sandholm2013:Population-Game}. Moreover, by using the
\rd as a carrier for diffusion, we actually obtain a special case of a
behavioral-strategy version of the replicator mutator dynamics. This
is important to know because it enables similar applications of
diffusion from perceptual confusion also in models of biological
evolution.

@@

This can matter to evolutionary dynamics. For illustration, take two
sender strategies whose behavior is different only at two
states. Compared to the second strategy, the first scores a much
higher expected utility at one state, but a slightly lower at the
other, so that the total expected utility of the first strategy is
higher. Under evolutionary dynamics that work on entire strategies the
first strategy will increase in proportion, relative to the
second. Consequently, a locally less efficient type of behavior can be
demoted because it is correlated with efficient behavior at another
choice point. Evolutionary dynamics working on behavioral strategies
do not allow for that. Behavioral adaptations occur independently at
different choice points.

As a result, dynamics on behavioral strategies are more coarse-grained
and not necessarily equivalent to dynamics on mixed strategies. An
advantage of dynamics in behavioral strategies is that they reduce
complexity, which helps with numerical simulations. They might also be
a more natural choice for cultural dynamics, e.g., when an agent
observes and imitates another agent's strategy: it may be more
plausible that agents adopt the use of a single word, not an entire
lexicon, so to speak. On the other hand, biological evolution seems
more naturally modeled in terms of mixed strategies: it is a full set
of behavioral disposition that is bequeathed to the next generation.

@@

  \caption{Examples of Nosofsky-similarity (non-normalized) and
    derived probability of state confusion (normalized) for
    $\card{\States} = 50$ and different
    values of imprecision $\impairment$.}
  \label{fig:NosofskySim}
\end{figure}

We further assume that the probability of confusing any two states
$\mystate{i}$ and $\mystate{j}$ is proportional to their perceived
similarity and therefore obtained by normalization:
\begin{align*}
  C_{ij} = \frac{\similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \impairment)} {\sum_j \similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \impairment)} \,.
\end{align*}
Confusion of states is then also a function of imprecision parameter
$\impairment$ (see Figure~\ref{fig:NosofskySim}). For $\impairment =
0$ the confusion matrix has $C_{ii} = 1$; for $\impairment > 0$, $C$
is positive, i.e., $C_{ij} >0$ for all $i$ and $j$; for $\impairment
\rightarrow \infty$ we find $C_{ij} = \nicefrac{1}{\card{\States}}$
for all $i$ and $j$.

@@

\subsection{Exploring the replicator diffusion dynamics}
\label{sec:simulations}

@@

These examples already show two interesting things. Firstly, inclusion
of imprecision can lead to seemingly well-behaved, yet vague
strategies in the sense that we are after. The sender strategy in
Figure~\ref{fig:example_stratsB} identifies clear positive and clear
negative cases for each signal, with a smooth transition
in-between. The receiver's interpretations of signals can be seen as
smoothed-out prototype regions. Secondly, (sender) strategies can
approach non-convex pure strategies under the replicator dynamic. We
see this in our limited-time simulations, but this also holds, for
some types of utility functions, for the limiting case. This was first
observed by Elliott Wagner in unpublished work
\citep[c.f.][]{OConnor2013:Evolving-Percep}. A full analysis of the
dynamics of sim-max games is beyond the scope of this paper, but we
will see shortly that diffusion from confusability of states clearly
prevents evolutionary paths that meander for a long time in the
vicinity of non-convex strategies.

@@

To further explore our simulation results, we calculated
metrics that aim to numerically capture how vague, generally
well-structured, and communicatively efficient the recorded
strategy pairs were. \emph{Entropy} captures the amount of
systematicity or regularity in signal use. \emph{Convexity} captures
whether a behavioral strategy would project onto a convex pure
strategy. \emph{Expected utility} measures the communicative
efficiency of evolved strategy pairs.\todo{take average of entropy}

@@

\paragraph{Convexity.} At least for sender strategies, which converge
faster, it also makes sense to define a categorical measure of
convexity that compensates for potential vagueness. To determine
whether a sender strategy $\Sstrat$ is convex despite possibly being
vague, we look at the derived pure strategy $\Spure$ for which
$\Spure(\state) = \argmax_{\messg' \in \Messgs}
\Sstrat(\state,\messg')$. If that $\Spure$ is convex, we also count
$\Sstrat$ as convex. The sender strategy in
Figure~\ref{fig:example_stratsA} is not convex, while the one in
Figure~\ref{fig:example_stratsB} is.  One would a priori not expect a
systematic relation between imprecision and convexity.

@@

As expected, increasing imprecision leads to higher entropy and lower
expected utility. Importantly, however, imprecision does not
necessarily lead to disastrous decline of communicative success. What was
not expected, on the other hand, was a relation between imprecision and
convexity. However, the corresponding plot clearly shows that
higher imprecision led to a higher number of outcomes with
convex sender strategies. Moreover, it also led to higher likelihood of
convergence. In fact, sufficient impairment always ensured convergence
and convexity. It appears that perceptual imprecision leads to
more entropy, slightly less communicative efficiency, but more
regular, well-behaved languages in shorter time.

@@

Interestingly, this type of variability in evolutionary outcomes can
be weeded out by impairment. To investigate this, we calculated the average
distance between evolved sender strategies within each group of trials
that had identical parameter values. We determined the distance
between strategies $\Sstrat$ and $\Sstrat'$ as the average Hellinger
distance between distributions $\Sstrat(\state)$ and
$\Sstrat'(\state)$ at each choice point $\state$:
\begin{align*}
  \text{HD}(\Sstrat,\Sstrat') & = \frac{1}{{\card{\States} \cdot
     \sqrt{2}}} \cdot  \sum_{\state \in \States} 
 \sqrt{\sum_{\messg \in  \Messgs}
         \left ( \sqrt{\Sstrat(\state,\messg)} -
         \sqrt{\Sstrat'(\state,\messg)} \right )^2 }\,.
\end{align*}
To compensate for the arbitrariness of message use, we set the
distance between strategies $\Sstrat$ and $\Sstrat'$ to be the maximum
of $\text{HD}(\Sstrat,\Sstrat')$ and $\text{HD}(\Sstrat^*,\Sstrat')$
where $\Sstrat^*$ is $\Sstrat$ with reversed message indices. An example of the
\emph{within group distances}, i.e., the average distances between all
sender strategies obtained for the same parameter values, is plotted
on the left in Figure~\ref{fig:GroupMeasures} for $\beta =
0.1$ and $n_s = 10$. Despite some quantitative differences, the
general trend is the same for all other parameter settings that we tested:
with increasing imprecision, the sender strategies evolving under \rdd
were much more alike (modulo swapping of messages). This means that
perceptual imprecision unifies evolutionary outcomes and guarantees
sender strategies that are not only convex, but also regular in that
they induce a vague category split exactly in the middle of the unit
interval.

@@

Admittedly, this argument hinges on the choice of utility
function. Still, to the extent that the utility functions we chose are
reasonable---and we think they are very reasonable---the case
suggests that quantal responses do not paint a complete picture. Naturally,
we cannot claim our approach provides valid predictions for all cases.
Given the complexity of the phenomenon, and the fact that
the quantal response approach is built on very reasonable assumptions,
it is likely that both proposals are relevant pieces of the puzzle.
%Vagueness is more
%plausibly an effect of perceptual confusion of similar states than of
%computational errors in maximizing expected utility.

>>>>

\begin{abstract}
  Signaling games are popular models for studying the evolution of
  meaning, but typical approaches do not incorporate vagueness as a
  feature of successful signaling.  Complementing recent like-minded
  models, we introduce the replicator diffusion dynamic as a special
  case of the replicator mutator dynamic that combines evolutionary
  pressure on optimal signal use with stochastic noise in the
  processing of similar stimuli. Applying this new dynamic to a
  generalization of Lewis' signaling games, we show that stochastic
  imprecision leads to vague, yet by-and-large communicative efficient
  signal use, and, moreover, that it unifies evolutionary outcomes and helps
  avoid suboptimal categorization.
\end{abstract}

@@

An alternative approach is taken by
\citet{FrankeJager2010:Vagueness-Signa} and
\citet{OConnor2013:The-Evolution-o} who show how the above desiderata
can be met by evolving strategies in sim-max games. The approach we
take here is similar in spirit, but different in relevant detail. We
argue here that \citeauthor{FrankeJager2010:Vagueness-Signa}'s
approach gives rise to problematic predictions under certain
conditions which are avoided in our approach. While
\citeauthor{OConnor2013:The-Evolution-o} gives an agent-level learning
dynamic under which vague signaling arises, we complement the picture
by giving a population-level dynamic that abstracts from the concrete
behavior of signalers. A more in-depth comparison is deferred until
Section~\ref{sec:discussion}. Let us first look into our own proposal
in more detail.

@@

The most straightforward interpretation of this sequential definition
is that the \rdd captures the most likely path of development of
aggregate behavior in a society, in which behavioral changes occur
independently at each choice point, as described by the \rd for
behavioral strategies. Whatever strategy an agent does actually hold,
its realization is perturbed by imperfect discriminability of states
in such a way that the expected overall aggregate behavior is the
diffusion of the strategies that the agents would play in a noise-free
world. Any agent-level interpretation of the \rd in behavioral
strategies seems \emph{prima facie} compatible with this
interpretation.

The sequential definition highlights that other (discrete-time)
evolutionary dynamics can be subjected to state-confusability in an
analogous fashion. It is easy, for instance, to perturb a
best-response dynamic in case we wanted to model cognizing, rational
agents with a noisy perceptual system. Our choice of the \rd as a
carrier dynamic is motivated by its generality and austerity, being
compatible with a variety of implicit agent-level behavior, without
presupposing extravagant higher cognitive abilities
\citep[c.f.][]{Sandholm2013:Population-Game}. Moreover, by using the
\rd as a carrier for diffusion, we actually obtain a special case of a
behavioral-strategy version of the replicator mutator dynamics. This
is important to know because it enables similar applications of
diffusion from perceptual confusion also in models of biological
evolution.

@@

This can matter to evolutionary dynamics. For illustration, take two
sender strategies whose behavior is different only at two
states. Compared to the second strategy, the first scores a much
higher expected utility at one state, but a slightly lower at the
other, so that the total expected utility of the first strategy is
higher. Under evolutionary dynamics that work on entire strategies the
first strategy will increase in proportion, relative to the
second. Consequently, a locally less efficient type of behavior can be
promoted because it is correlated with efficient behavior at another
choice point. Evolutionary dynamics working on behavioral strategies
do not allow for that. Behavioral adaptations occur independently at
different choice points.

As a result, dynamics on behavioral strategies are more coarse-grained
and not necessarily equivalent to dynamics on mixed strategies. An
advantage of dynamics in behavioral strategies is that they reduce
complexity, which helps with numerical simulations. They might also be
a more natural choice for cultural dynamics, e.g., when an agent
observes and imitates another agent's strategy: it may be more
plausible that agents adopt the use of a single word, not an entire
lexicon, so to speak. On the other hand, biological evolution seems
more naturally modeled in terms of mixed strategies, when it is a full
set of behavioral disposition that is bequeathed to the next
generation.

@@

  \caption{Examples of Nosofsky-similarity for different values of
    imprecision.}
  \label{fig:NosofskySim}
\end{figure}

We further assume that the probability of confusing any two states
$\mystate{i}$ and $\mystate{j}$ is proportional to their perceived
similarity and therefore obtained by normalization:
\begin{align*}
  C_{ij} = \frac{\similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \impairment)} {\sum_j \similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \impairment)} \,.
\end{align*}
Confusion of states is then also a function of imprecision parameter
$\impairment$. For $\impairment = 0$, the confusion matrix has $C_{ii}
= 1$; for $\impairment > 0$, $C$ is positive, i.e., $C_{ij} >0$ for
all $i$ and $j$; for $\impairment \rightarrow \infty$, we find $C_{ij}
= \nicefrac{1}{\card{\States}}$ for all $i$ and $j$.

@@

\subsection{Numerical simulations}
\label{sec:simulations}

@@

These examples already show two interesting things. Firstly, inclusion
of imprecision can lead to seemingly well-behaved, yet vague
strategies in the sense that we are after. The sender strategy in
Figure~\ref{fig:example_stratsB} identifies clear positive and clear
negative cases for each signal, with a smooth transition
in-between. The receiver's interpretations of signals can be seen as
smoothed-out prototype regions. Secondly, (sender) strategies can
approach non-convex pure strategies under the replicator dynamic and
linger there for vast amounts of time, possibly indefinitely. We see
this in our limited-time simulations, but this also holds, for some
types of utility functions, for the limiting case. This was first
observed by Elliott Wagner in unpublished work
\citep[c.f.][]{OConnor2013:Evolving-Percep}. A full analysis of the
dynamics of sim-max games is beyond the scope of this paper, but we
will see shortly that diffusion from confusability of states clearly
prevents evolutionary paths that meander for a long time in the
vicinity of non-convex strategies.

@@

To further explore our simulation results, we calculated metrics that
aim to numerically capture how vague, generally well-structured, and
communicatively efficient the recorded strategy pairs
were. \emph{Entropy} captures the amount of systematicity or
regularity in signal use. \emph{Convexity} captures whether a
behavioral strategy would project onto a convex pure
strategy. \emph{Expected utility} measures the communicative
efficiency of evolved strategy pairs.

@@

\paragraph{Convexity.} At least for sender strategies, which develop
faster than receiver strategies, it also makes sense to define a
categorical measure of convexity that compensates for potential
vagueness. To determine whether a sender strategy $\Sstrat$ is convex
despite possibly being vague, we look at the derived pure strategy
$\Spure$ for which $\Spure(\state) = \argmax_{\messg' \in \Messgs}
\Sstrat(\state,\messg')$. If that $\Spure$ is convex, we also count
$\Sstrat$ as convex. The sender strategy in
Figure~\ref{fig:example_stratsA} is not convex, while the one in
Figure~\ref{fig:example_stratsB} is.  One would a priori not expect a
systematic relation between imprecision and convexity.

@@

As expected, increasing imprecision leads to higher entropy and lower
expected utility. Importantly, however, imprecision does not
necessarily lead to disastrous decline of communicative success. What
was not expected, on the other hand, was a relation between
imprecision and convexity. However, the corresponding plot clearly
shows that higher imprecision led to a higher number of outcomes with
convex sender strategies. Moreover, it also led to higher likelihood
of convergence. In fact, sufficient impairment always ensured
convergence and convexity. It appears that perceptual imprecision
leads to more vagueness, slightly less communicative efficiency, but
more regular, well-behaved languages in shorter time.

@@

Interestingly, this type of variability in evolutionary outcomes can
be weeded out by impairment. To investigate this, we calculated the average
distance between evolved sender strategies within each group of trials
that had identical parameter values. We determined the distance
between strategies $\Sstrat$ and $\Sstrat'$ as the average Hellinger
distance between distributions $\Sstrat(\state)$ and
$\Sstrat'(\state)$ at each choice point $\state$:
\begin{align*}
  \text{HD}(\Sstrat,\Sstrat') & = \frac{1}{{\card{\States} \cdot
     \sqrt{2}}} \cdot  \sum_{\state \in \States} 
 \sqrt{\sum_{\messg \in  \Messgs}
         \left ( \sqrt{\Sstrat(\state,\messg)} -
         \sqrt{\Sstrat'(\state,\messg)} \right )^2 }\,.
\end{align*}
To compensate for the arbitrariness of message use, we set the
distance between strategies $\Sstrat$ and $\Sstrat'$ to be the maximum
of $\text{HD}(\Sstrat,\Sstrat')$ and $\text{HD}(\Sstrat^*,\Sstrat')$
where $\Sstrat^*$ is $\Sstrat$ with reversed message indices. An example of the
\emph{within group distance}, i.e., the average distances between all
sender strategies obtained for the same parameter values, is plotted
on the left in Figure~\ref{fig:GroupMeasures} for $\beta =
0.1$ and $n_s = 10$. Despite some quantitative differences, the
general trend is the same for all other parameter settings that we tested:
with increasing imprecision, the sender strategies evolving under \rdd
were much more alike (modulo swapping of messages). This means that
perceptual imprecision unifies evolutionary outcomes and guarantees
sender strategies that are not only convex, but also regular in that
they induce a vague category split exactly in the middle of the unit
interval.

@@

Admittedly, this argument hinges on the choice of utility
function. Still, to the extent that the utility functions we chose are
reasonable---and we think they are very reasonable--- the case
suggests that quantal responses alone are not a good model,
intuitively speaking, for why linguistic categories are
vague. Vagueness is more plausibly an effect of perceptual confusion
of similar states than of computational errors in maximizing expected
utility. Even more plausibly, though, perceptual confusion \emph{and}
computational errors, together with even further factors such as
limited observations, imperfect memory and more, are needed for a
complete understanding of why vague categorization and linguistic
meaning as a natural concomitant of a distinguished cognitive
make-up.
